The evolution of gene expression in the term placenta of viviparous mammals

### 2016-02-22 presentation/placenta_viviparous_presentation.Rnw
fix gene trees for most of the plots

\begin{frame}{Primate Placenta Transcriptome}
\begin{table}
  \footnotesize
<<primate.lineage.specific,echo=FALSE,results="asis">>=
pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
setkey(pri.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed significantly in the primate
  order as compared to all other analyzed species.}
\label{tab:primate_lineage_specific}
\end{table}
\end{frame}

@@

% \begin{frame}{Prolactin}
% \begin{itemize}
% \item \elephant is only analyzed mammal with high prolactin expression
%   (\textit{PRL}) in the placenta
% \item However, \mouse, \carmeli, and \cow have duplications of
%   \textit{PRL} with expression}
% \end{itemize}
% 
% \begin{figure}
% <<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
% 
% tree.id <-
%     genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
% gene.tree.table <- trees[[tree.id]]$table
% gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
% gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
% plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
%                  gene.tree.table,gene.tree.expression,
%                  axis.text.size=5,
%                  fontsize=5)
% @ 
% \end{figure}
% 
% \end{frame}

@@

\begin{frame}{CGB}
\begin{figure}
<<chorionic.somatomammotropin.beta.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 subtree=c("ENSMODG00000013768","ENSG00000104818"),
                 offset.ratio=1,
                 fontsize=plot.label.size,
                 axis.text.size=3.5
                 )
@ 
\end{figure}
\end{frame}

@@

\begin{frame}
Relaxin (\textit{RLN}) is a hormone that has multiple effects during
pregnancy including increasing maternal blood flow, widening the pubic
bone and relaxing uterine musculature in preparation for labor. In
\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
of relaxin in the term placenta, with little evidence in other species
examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
to be produced in the placenta of the pregnant \dog and is used in the
diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
the luteal ovary and placenta of \pig
\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
source of relaxin produced by pregnant \elephant remains unclear
\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
\elephant has been suggested to be a major signal for the initiation
of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
Here we show that the placenta of elephants show high expression of
relaxin and is likely a major source of \textit{RLN} at term. In
contrast to the results of
\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
of \textit{RLN} in the placenta of \mouse and \carmeli.

@@

\begin{frame}{Galectins}
Galectins are recognized as major regulators of pregnancy in both
fetal and maternal tissues
\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
evolution of new lineage-specific galectins has occurred in both
primates and ruminants, with the genes \textit{LGALS13},
\textit{LGALS14}, and \textit{LGALS16} recognized as being
placenta-specific in \human and regulating immune responses by
inducing apoptosis of maternal T-cells
\cite{Than.ea2009:primatesubfamilyofgalectins} (Supplemental
\cref{fig:galectins_lgals13_evolution_tree}). Indeed, \textit{LGALS14}
has the highest expression of any galectin in the \human and \smonkey
placenta (\cref{fig:hormone_genes}, Supplemental
\cref{fig:galectins_lgals13_evolution_tree}), and other primate
galectins such as \textit{LGALS13} and \textit{LGALS16} are also
expressed in \human and \smonkey placenta. However, in the closely
related \bonobo, \textit{LGALS14} is not found, and \textit{LGALS1} is
the most abundant galectin. \textit{LGALS1} shows generally high
expression in all placentas analyzed. \textit{LGALS1} is also involved
in mediating maternal-fetal immune tolerance, and is expressed in both
maternal and fetal tissue in primates
\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
expression was thought to be confined to eutherian mammals, as there
is evidence of adaptive evolution at this node, as well as the
appearance of estrogen response elements in eutherian cis-regulatory
regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
to whether galectins such as \textit{LGALS1} were important in the
regulation of pregnancy in marsupials, as well as eutherians. Here we
show substantial expression of galectins in the opossum, especially
\textit{LGALS1} (Supplemental \cref{fig:galectins_evolution_tree}). In
addition, \textit{LGALS2} is also expressed. The presence of galectins
at the maternal-fetal interface in \opossum implies that it was also
present in the ancestral placenta of therian mammals. Indeed, adaptive
evolution of \textit{LGALS1} on the eutherian stem is likely due to
the increased invasive nature of eutehrian placentation rather than
the origination of maternal-fetal tolerance. Although the function of
\textit{LGALS1} in the marsupial placenta needs to be confirmed, it is
likely that the mediation of maternal-fetal immune tolerance by
galectins is an ancient feature of mammalian viviparity.

@@

\begin{frame}
The \opossum placenta shows expression of genes that either do not
exist in eutherians or are not found in the term eutherian placenta
(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
of hemoglobin in fetal tissues at term, as evidenced by the expression
of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
generally due to their very short gestation periods. Embryonic
hemoglobins in marsupials have a lower affinity for oxygen and allow
for a lower rate of oxygen uptake from maternal tissues
\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
found in eutherians; Ensembl gene trees reveal that homologs of this
gene exist in other amniotes such as birds and turtles, but are not
present in eutherian mammals. The homolog of this gene in chicken is
found to expressed in reproductive tract
\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
family and involved in innate immunity. It is expressed in the mammary
gland of the tammar wallaby during pregnancy and early lactation
\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
relative abundance in \opossum placenta.

@@

\begin{frame}{PRSS16}

11 genes that are related
to the \textit{PRSS16} gene, a thymus-specific serine protease that
may be involved regulating the presentation of self-peptides in CD4+
T-lymphocytes \cite{Fornari.ea2012:TranscriptionprofilingofPrss16}.
All 11 of these genes are expressed in the elephant placenta, three of
which have very abundant mRNAs
(\cref{tab:top50_elephant_placenta} and \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}C). It is
possible that this elephant-specific gene family may be involved in
immunity in the elephant placenta and the mediation of maternal-fetal
immune tolerance; however, future work will need to test these
hypotheses. 
\begin{figure}
<<prss16.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
                 fontsize=4,
                 width=3,
                 axis.text.size=axis.text.size,
                 offset.ratio=1.2
                 )

@ 
\end{figure}
\end{frame}

@@

\item Identified common expression patterns of

@@

@@

\begin{frame}[plain]{References}
  \begin{center}
    \mbox{}\vspace{-\baselineskip}
    \printbibliography[heading=none]
  \end{center}
\end{frame}

>>>>

\begin{frame}{Primate Placenta Transcriptome}
\begin{table}
  \footnotesize
<<primate.lineage.specific,echo=FALSE,results="asis">>=
pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
setkey(pri.all.lin.spec,"Human Symbol")
@ 
\end{table}
\end{frame}

@@

\begin{frame}{Prolactin}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item \elephant is only analyzed mammal with high prolactin expression
  (\textit{PRL}) in the placenta
\item However, \mouse, \carmeli, and \cow have duplications of
  \textit{PRL} with expression
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,
                 axis.text.size=5,
                 fontsize=4)
@ 
\end{figure}
\end{columns}
\end{frame}

@@

\begin{frame}{CGB}
  \begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item \textit{CGB} is highly expressed in \human

\item Some expansion of \textit{CGB} in \bonobo, but not highly expressed
\item High expression of \textit{LHB} in \smonkey
\begin{itemize}
\item May be indicative of expansion
\item Would have to examine assembled transcripts to tell
\end{itemize}
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<chorionic.somatomammotropin.beta.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 subtree=c("ENSMODG00000013768","ENSG00000104818"),
                 offset.ratio=1,
                 fontsize=10,
                 axis.text.size=3.5
                 )
@ 
\end{figure}
\end{columns}
\end{frame}

@@

\begin{frame}{Relaxin}
  \begin{columns}
    \column{0.4\textwidth}
    \begin{itemize}
    \item \textit{RLN} increasing maternal blood flow, widening the pubic
      bone and relaxing uterine musculature in preparation for labor
    \item We show that the placenta is a source of \textit{RLN} in \dog,
      \elephant, \mouse, \carmeli, and \horse.
    \end{itemize}
    \column{0.6\textwidth}
    \begin{figure}
<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,
                 axis.text.size=5,
                 fontsize=5)
@ 
\end{figure}
\end{columns}
\end{frame}

@@

\begin{frame}{Galectins}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item Galectins are major regulators of pregnancy in fetal and
  maternal tissues
\item \textit{LGALS1} mediates maternal-fetal
  immune tolerance
\item substantial expression of galectins in the \opossum
\item consistent with ancestral placenta may have maternal-fetal
  immune tolerance mediated by galectins
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,
                 offset=0.8,
                 axis.text.size=7,
                 fontsize=7,
                 subtree=c("ENSMODG00000000457","ENSCAFG00000016577")
                 )
@ 
\end{figure}
\end{columns}
\end{frame}

\begin{frame}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item Evolution of new lineage-specific galectins has occurred in both
  primates and ruminants, with the genes \textit{LGALS13},
  \textit{LGALS14}, and \textit{LGALS16} recognized as being
  placenta-specific in \human and regulating immune responses by
  inducing apoptosis of maternal T-cells
  \cite{Than.ea2009:primatesubfamilyofgalectins}
\end{itemize}

\column{0.6\textwidth}
\begin{figure}
<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,
                 offset=0.8,
                 axis.text.size=7,
                 fontsize=10
                 )
@ 
\end{figure}
\end{columns}
\end{frame}

@@

\begin{frame}
\begin{itemize}
\item The \opossum placenta shows expression of genes that either do
  not exist in eutherians or are not found in the term eutherian
  placenta

\item The existence of embryonic forms of hemoglobin in fetal tissues
  at term, as evidenced by the expression of \textit{HBE} and
  \textit{HBZ} is a feature of marsupials that is generally due to
  their very short gestation periods.

\item Embryonic hemoglobins in marsupials have a lower affinity for
  oxygen and allow for a lower rate of oxygen uptake from maternal
  tissues \cite{Holland.ea1997:Characterizationofembryonicglobin}.

\item One novel uncharacterized gene that shows high expression
  (\textit{LOC10001195}) is not found in eutherians; Ensembl gene
  trees reveal that homologs of this gene exist in other amniotes such
  as birds and turtles, but are not present in eutherian mammals.

\item \textit{WFDC2} is a member of the whey acidic protein (WAP)
  family and involved in innate immunity.
\end{itemize}
\end{frame}

@@

\begin{frame}{PRSS16}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item \textit{PRSS16}, a thymus-specific serine protease that
  regulates the presentation of self-peptides in CD4+ T-lymphocytes
  \cite{Fornari.ea2012:TranscriptionprofilingofPrss16}.
\item Possibly involved in the mediation of maternal-fetal immune
  tolerance in \elephant
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<prss16.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
                 fontsize=4,
                 width=3,
                 axis.text.size=axis.text.size,
                 offset.ratio=1.2
                 )

@@

\item Identified common expression patterns of the mammalian
  expression

@@

\item Lots to learn about what is actually driving the changes in
  expression

@@

\begin{frame}[plain,allowframebreaks]{References}
  \begin{center}
    \mbox{}\vspace{-\baselineskip}
    \printbibliography[heading=none]
  \end{center}
\end{frame}
